Handling String-Number Conversion in String Constraint Solving

### 2019-11-14 draft.tex
some  minor update

We introduce \emph{parametric flat automata} (PFA) that will be used to define patterns 
%for the string variable domain restriction. 
used by the under-approximation module to restrict the domain of string variables. 
Namely, an PFA is a finite automaton $A = (Q,T,\cvar,\st{0}{0},\st{m}{0})$ operating over an alphabet $\cvar$ of \emph{character variables}, 
and satisfying the following structural constraints:
\begin{enumerate}
	\item The final state $\st{m}{0}$ is reached from  the initial state $\st{0}{0}$ through a straight path of $m-1$ transitions $(\st{i}{0},v_i,\st{i+1}{0}) \in T$, $\st{i}{0} \in Q$ for $i\in[0,m-1]$. %, with pairwise distiand states $\st{i}{0} \in Q$ for $i\in[0,m-1]$.
	\item  
Each state $\st{i}{0}$ is the origin of a unique simple cycle of the length $l_i\in\mathbb{N}$, consisting of states $\st{i}{j-1} \in Q$ and transitions $(\st{i}{j-1}, \sym{i}{j-1}, \st{i}{j \bmod l_i})$ for $j\in [1,l_i]$. 
Notice that case when $l_i = 0$ is also admissible and means that there is no cycle on $q_i$.
%Assume that $\st{i}{0}$ is included in a cycle with $l_i$ states (includes $\st{i}{0}$). 
%For each $j\in [1,l_i]$, we have $\st{i}{j} \in Q$ and $(\st{i}{j-1}, \sym{i}{j-1}, \st{i}{j \bmod l_i})$. 
%(where $l_i = 0$ means that there is no cycle on $q_i$)
	\item Each variable in $\cvar$ appears on at most one transition.% (but $\epsilon$ transition can appear multiple times). 
\end{enumerate} 
\begin{figure*}
	\tikzset{state/.style={circle,draw=blue!50,fill=blue!20,
			thick,inner sep=0pt,minimum size=6mm}, initial text=$ $}
	\begin{tikzpicture} 
		\node[state,initial] (q0)  {$\st{0}{0}$};
		\node[state] (q1) [right = 3cm of q0] {$\st{1}{0}$};
		\node (q2) [right = 1.8cm of q1]{};
		\node (q3) [right = 1cm of q2]{};
		\node[state,accepting] (qm) [right = 1.8cm of q3] {$\st{m}{0}$};

@@

\paragraph{Synchronization of PFA}
For two PFA $A$ and $A'$ over character variables $V$ and $V'$, respectively, 
it builds \emph{synchronization formula} $\syncfof A {A'}$, a linear formula over $\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo$ encoding the semantic intersection $\semof A \cap \semof {A'}$. The synchronization formula will be used in, e.g., the decision procedure of equality constraints $t_1=t_2$ to find a common interpretation of $t_1$ and $t_2$. More precisely, the models of the synchronization formula $\syncfof A {A'}$ correspond to the set of all pairs of word encodings $(I\cup \pim)\in \parikhof {A}$ and $(I'\cup \pim')\in\parikhof {A'}$ such that 
$\decode A(I\cup \pim) = \decode {A'}(I'\cup \pim')$, which constitute the intersection.

Particularly, a synchronization formula is build for two PFA  
$A = (Q,T,\cvarone,q_i,q_f)$ and ${A'} = (Q',T',\cvartwo,q_i',q_f')$ such that $\cvarone \cap \cvartwo = \emptyset$. 
%
It is extracted from the \emph{asynchronous product} of $A$ and $A'$. 
%
The asynchronous product is an automaton that uses $Q\times Q'$ as the set of states and 
$\cvarone_\epsilon\times\cvartwo_\epsilon$
as the alphabet. 
%
Every accepting run of $\syncof A {A'}$ corresponds to a pair of accepting runs, 
a run of $A$ over a parametric word $x$ and a run of ${A'}$ over a parametric word $x'$. 
The word accepted by the run of $\syncof A {A'}$ induces constraints on the interpretations of $I$ of $\cvarone$ and $I'$ of $\cvartwo$ under which the two parametric words have the same interpretation, i.e. $I(x) = I(x')$.

@@

The correctness of this construction is stated in Lemma~\ref{lemma:synccorrect} below. The lemma can also be understood as a precise specification the semantics of the synchronization formula, on which we rely for correctness of the construction of under-approximations of string constraints. 
%
\begin{lemma}\label{lemma:synccorrect}
%$\semof {\exists V_\times . \syncfof A {A'}}\restrict{\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
$\semof {\syncfof A {A'}}_{\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
\end{lemma}
%
Another version:
\begin{lemma}
$
\semof {\syncfof A {A'}}_{\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo} = 
\{
(I\cup\parikhof{x})\cup(I'\cup\parikhof{x'}) \mid 
x\in L(A), x'\in L(A'), I(x) = I(x')
\}
$
\end{lemma}
Notice that if $A$ is flat (or, symmetrically, if $A'$ is flat), then $\semof A \cap \semof {A'}$ can be still decoded from $\semof {\syncfof A {A'}}$ due to Corollary~\ref{corollary:pfa}, 
namely in the form of\\ $\semof A \cap \semof {A'} = \{\decode A(I\cup x) \mid \exists J.((I\cup x) \cup J) \in \semof{\syncfof A {A'}}\}$.  
\bigskip

@@

@@

\yfc{The version here is buggy, but fixable. I am not sure what is the best fix, so for now I just leave it here. The problem is we cannot synchronize a PFA and a general parametric automaton (which can be non-flat)}
Assume that the constraint we take from $\Phi_I$ is a regular constraint $\phi_r \defeq x \in L(A)$. Assume that $A=(Q,T,\Sigma,q_i,q_f)$. First we create a PFA $A' = (Q,T', V',q_i,q_f)$ that has exactly the same set of states as $A$. The set of transitions $T'$ and character variables $V'$ is defined as follows:

\begin{itemize}
	\item $V' = \{ v_{(q,a,q')} \mid (q,a,q')\in T\}$
	\item $T' = \{ (q, v_{(q,a,q'), q') } \mid (q,a,q')\in T\}$
\end{itemize}
We use the linear formula $\phi_c \defeq \bigwedge_{(q,a,q')\in T} {v_{(q,a,q')} = a}$ to remember the relation between the fresh character variables in $V'$ and the symbol it represents. The relation between $A$ and $A'$ is expressed in Lemma~\ref{lemma:fa_pfa}.

\begin{lemma}\label{lemma:fa_pfa}
	$L(A) = ( {\semof {\syncfof {A_x }{A'}} \cap \bigcap_{\phi\in\Phi_C} \semof {\phi} } )_{V_x \cup \#V_x} = \{(I\cup\pim) \mid \decode {A_x}(I\cup\pim) \in L(A)\}$. 
\end{lemma}

@@

\paragraph{PFA that allows efficient string-integer conversion: } In the previous discussion, we know that it is unlikely to use arbitrary PFA and gain efficient solving procedure, due to the requirement to solve exponential integer constraints. The problem is that each loop in the PFA creates an exponential component. An immediate idea would be that maybe we can restrict the variable to some specific PFA that allow more efficient constraint solving. At the same time, we want to keep some good property, i.e., the search space covers all integers with at most $m$-digits.

>>>>

We introduce \emph{parametric flat automata} (PFA) that will be used to define patterns 
used by the under-approximation module to restrict the domain of string variables. 
A \emph{parametric automaton} (PA) is a finite automaton operating over an alphabet $\cvar$ of \emph{character variables}. 
A parametric flat automaton is a parametric automaton $A = (Q,T,\cvar,\st{0}{0},\st{m}{0})$ satisfying the following structural constraints:
\begin{enumerate}
	\item The final state $\st{m}{0}$ is reached from  the initial state $\st{0}{0}$ through a straight path of $m-1$ transitions $(\st{i}{0},v_i,\st{i+1}{0}) \in T$, $\st{i}{0} \in Q$ for $i\in[0,m-1]$. %, with pairwise distiand states $\st{i}{0} \in Q$ for $i\in[0,m-1]$.
	\item  
Each state $\st{i}{0}$ is the origin of a unique simple cycle of the length $l_i\in\mathbb{N}$, consisting of states $\st{i}{j-1} \in Q$ and transitions $(\st{i}{j-1}, \sym{i}{j-1}, \st{i}{j \bmod l_i})$ for $j\in [1,l_i]$. 
Notice that case when $l_i = 0$ is also admissible and means that there is no cycle on $q_i$.
%Assume that $\st{i}{0}$ is included in a cycle with $l_i$ states (includes $\st{i}{0}$). 
%For each $j\in [1,l_i]$, we have $\st{i}{j} \in Q$ and $(\st{i}{j-1}, \sym{i}{j-1}, \st{i}{j \bmod l_i})$. 
%(where $l_i = 0$ means that there is no cycle on $q_i$)
	\item Each variable in $\cvar$ appears on at most one transition.% (but $\epsilon$ transition can appear multiple times). 
\end{enumerate} 
\begin{figure*}
	\tikzset{state/.style={circle,draw=blue!50,fill=blue!20,
			thick,inner sep=0pt,minimum size=6mm}, initial text=$ $}
	\begin{tikzpicture} 
		\node[state,initial] (q0)  {$\st{0}{0}$};
		\node[state] (q1) [right = 3cm of q0] {$\st{1}{0}$};
		\node (q2) [right = 1.8cm of q1]{};
		\node (q3) [right = 1cm of q2]{};
		\node[state,accepting] (qm) [right = 1.8cm of q3] {$\st{m}{0}$};

@@

\paragraph{Synchronization of PAs}
For two PAs $A$ and $A'$ over character variables $V$ and $V'$, respectively, 
it builds \emph{synchronization formula} $\syncfof A {A'}$, a linear formula over $\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo$ encoding the semantic intersection $\semof A \cap \semof {A'}$. The synchronization formula will be used in, e.g., the decision procedure of equality constraints $t_1=t_2$ to find a common interpretation of $t_1$ and $t_2$. More precisely, the models of the synchronization formula $\syncfof A {A'}$ correspond to the set of all pairs of word encodings $(I\cup \pim)\in \parikhof {A}$ and $(I'\cup \pim')\in\parikhof {A'}$ such that 
$\decode A(I\cup \pim) = \decode {A'}(I'\cup \pim')$, which constitute the intersection.

Particularly, a synchronization formula is build for two PAs  
$A = (Q,T,\cvarone,q_i,q_f)$ and ${A'} = (Q',T',\cvartwo,q_i',q_f')$ such that $\cvarone \cap \cvartwo = \emptyset$. 
%
It is extracted from the \emph{asynchronous product} of $A$ and $A'$. 
%
The asynchronous product is an automaton that uses $Q\times Q'$ as the set of states and 
$\cvarone_\epsilon\times\cvartwo_\epsilon$
as the alphabet. 
%
Every accepting run of $\syncof A {A'}$ corresponds to a pair of accepting runs, 
a run of $A$ over a parametric word $x$ and a run of ${A'}$ over a parametric word $x'$. 
The word accepted by the run of $\syncof A {A'}$ induces constraints on the interpretations of $I$ of $\cvarone$ and $I'$ of $\cvartwo$ under which the two parametric words have the same interpretation, i.e. $I(x) = I(x')$.

@@

The correctness of this construction is stated in Lemma~\ref{lemma:synccorrect} below. The lemma can also be understood as a precise specification the semantics of the synchronization formula, on which we rely for correctness of the construction of under-approximations of string constraints. 
%
\hide{
\begin{lemma}\label{lemma:synccorrect}
%$\semof {\exists V_\times . \syncfof A {A'}}\restrict{\cvarone \cup \cvartwo \cup \pvarsof{\cvarone} \cup \pvarsof\cvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
$\semof {\syncfof A {A'}}_{\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo} = \{(I\cup\pim)\cup(I'\cup\pim') \mid \decode A(I\cup\pim) = \decode {A'}(I'\cup\pim')\}$. 
\end{lemma}
%
Another version:}
\begin{lemma}\label{lemma:synccorrect}
$
\semof {\syncfof A {A'}}_{\cvarone\cup \cvartwo \cup \pvarone\cup\pvartwo} = 
\{
(I\cup\parikhof{x})\cup(I'\cup\parikhof{x'}) \mid 
x\in L(A), x'\in L(A'), I(x) = I(x')
\}
$
\end{lemma}
Notice that if $A$ is flat (or, symmetrically, if $A'$ is flat), then $\semof A \cap \semof {A'}$ can be still decoded from $\semof {\syncfof A {A'}}$ due to Corollary~\ref{corollary:pfa}, 
namely in the form of $\semof A \cap \semof {A'} = \{\decode A(I\cup \pim) \mid (I\cup \pim) \in \semof{\syncfof A {A'}}_{\cvarone\cup \pvarone}\}$.

@@

Assume that the constraint we take from $\Phi_I$ is a regular constraint $\phi_r \defeq x \in L(A)$. Assume that $A=(Q,T,\Sigma,q_i,q_f)$. First we create a PA $A' = (Q,T', V',q_i,q_f)$ that has exactly the same set of states as $A$. The set of transitions $T'$ and character variables $V'$ is defined as follows:

\begin{itemize}
	\item $V' = \{ v_{(q,a,q')} \mid (q,a,q')\in T\}$
	\item $T' = \{ (q, v_{(q,a,q'), q') } \mid (q,a,q')\in T\}$
\end{itemize}
We use the linear formula $\phi_c \defeq \bigwedge\limits_{(q,a,q')\in T} (v_{(q,a,q')} = a)$ to remember the relation between the fresh character variables in $V'$ and the symbol it represents. The relation between $A$ and $A'$ is expressed in Lemma~\ref{lemma:fa_pfa}.

\begin{lemma}\label{lemma:fa_pfa}
	$L(A) =  \{  \decode {A'}(I\cup\pim) \mid (I\cup\pim)\in  {\semof {\parikhfof {A'} \wedge \phi_c}} _{V' \cup \#V'} \}$. 
\end{lemma}

@@

@@

\paragraph{PFA that allows efficient string-number conversion: } In the previous discussion, we know that it is unlikely to use arbitrary PFA and gain efficient solving procedure, due to the requirement to solve exponential integer constraints. The problem is that each loop in the PFA creates an exponential component. An immediate idea would be that maybe we can restrict the variable to some specific PFA that allow more efficient constraint solving. At the same time, we want to keep some good property, i.e., the search space covers all integers with at most $m$-digits.
